Moisture absorption and drying alter nonadditive litter flammability in a mixed conifer forest

### 2020-04-29 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Edits in response to reviews

\documentclass[letterpaper,12pt]{article}
\usepackage{jecol}

@@

      \item To fill these gaps in knowledge, we determined moisture absorption
    and retention characteristics of the litter from multiple tree species,
    investigated what litter traits influenced these processes, and tested for
    nonadditive moisture dynamics and nonadditive flammability in multi-species
    mixtures. The litter from eight tree species of a mixed-conifer forest was
    saturated and allowed to dry to determine moisture absorption capacity and
    drying rates. Burn trials were performed in litter beds of single species
    and of mixtures to establish flammability response to the dry-down process.

      \item Moisture dynamics varied across species with higher specific leaf
    area being associated with higher maximum absorption and lower litter bulk
    density being associated with faster drying rates. We examined two axes of
    flammability variation and found that these differing moisture dynamics
    resulted in time since wetting having strongly differing effects on
    flammability across the major litter types typified by different tree taxa.

      \item Litter mixtures exhibited nonadditivity in flame spread rate and
    in fuel consumption. This nonadditivity became more positive with fuel
    drying.

@@

Fuel moisture content is an important, and often driving, determinant of
surface fire behavior \citep{Rothermel-1972}. Moisture acts as a heat-sink
resulting in less energy available for propagation, and the resulting water
vapor will both cool the flaming front and dilute flammable gases
\citep{Albini-1976, Shafizadeh-1977}. Environmental conditions such as
temperature, relative humidity, and wind drive the moisture content of the fuel
available to burn \citep{Kreye+Varner+etal-2018}. Variation in moisture of
living fuels has strong effects on flammability, but moisture of dead excised
litter fuels can also vary. In particular, fine dead fuels respond very readily
to environmental changes \citep{Nelson-2001}. Moisture of litter fuels varies
with climate, with seasons, and with short term weather within a season.
Therefore, both the rate of drying and flammability as a function of moisture
content are key parameters that influence fuel flammability throughout the fire
season.

Flammability is a multi-dimensional characteristic of fuels
\citep{Schwilk-2015, Pausas+Keeley+etal-2017} with different traits influencing
different components of fire behavior. In many forests that experience surface
fire, litter properties influence fire behavior
\citep{Ganteaume+Marielle+etal-2011, Schwilk+Caprio-2011,
  Varner+Kane+etal-2015}. In litter driven fire, the two major axes of
variation are described by measurements related to flame spread rate and
measurements related to heat release or heat transfer
\citep{Magalhaes+Schwilk-2012, Prior+Murphy+etal-2018}. These different
components of fire behavior have differing ecological effects with spread rate
being associated with rates of heat release and area burned, and with heat
transfer predicting soil impacts of fire and tree mortality. A shortcoming in
our current knowledge is our integration of specific species and plant traits
into fuel hazard assessment, fire modeling, and fire effects prediction.
Developing quantitative yet robust and general ways to integrate actual plant
communities into fuel assessment will require understanding the magnitude of
flammability variability across species in space and time and parameterizing
based on plant characteristics.

@@

Fuel moisture content through time is controlled by two processes, wetting
(including absorption and adsorption) and drying rate (desorbtion)
\citep{Anderson-1990, Kreye+Kobziar+etal-2013}. Absorptive capacity represents
the potential maximum fuel moisture and is evaluated by the moisture content of
litter when saturated; hereafter we refer to this measurement as ``maximum
water absorption''. Differences in moisture content over time since saturation
can indicate different abilities to retain moisture. Leaf morphology,
especially leaf cuticle thickness, can influence water absorption
\citep{Van_Wagner-1969}. Litter particle geometry may have effects on maximum
moisture content because a larger total surface area will increase water
adsorption. Drying rate should also be influenced by leaf and litter bed traits
because surface-area-to-volume ratio and litter bulk density will influence air
flow through the litter bed. Therefore, we expect litter packing to effect fire
behavior in two ways: directly by its effect on oxygen availability for
combustion, and indirectly by its effect on drying rate. Both these pathways
are affected by leaf particle traits, especially leaf size
\citep{Scarff+Westoby-2006}. Fuel-beds composed of larger leaves or leaves with
thin cuticles should favor moisture absorption. On the other hand, large leaves
lower litter density and should favor drying, thus increasing the probability
of ignition and fire spread. The interaction of moisture absorption and drying
deserves investigation.

In mixed species communities, species traits may interact such that the
flammability of multi-species mixtures is not the average of the constituent
species. Such nonadditive effects are often synergistic in dry litter fuels
where the flammability of a mixture is driven by the most flammable species
\citep{VanAltena+Logtestjin+etal-2012, Magalhaes+Schwilk-2012}, but negative
nonadditivity has also been recorded for litter mixtures
\citep{Blauw+Wensink+etal-2015, Zhao+vanLogtestijn+etal-2019}. It is possible
that similar nonadditive effects may influence litter moisture content and
flammability of moist fuels. There is some evidence for nonadditive
flammability in moist fuels with one study finding increasing nonadditivity
with moisture \citep{Blauw+Wensink+etal-2015}, but this previous work did not
include estimation of the dry-down process and had an unclear relationship to
fire behavior int he field.. In mixed conifer forest in California, where the
fire regime is characterized by surface fires fueled by plant litter, overstory
tree composition drives variation in local fire behavior
\citep{Schwilk+Caprio-2011} as a result of differential litter flammability
across tree species and as a result of synergistic interactions in
multi-species litter mixtures \citep{Magalhaes+Schwilk-2012}. These
flammability feedbacks may help explain historical fire frequency patterns
\citep{Schwilk+Caprio-2011}, but current estimates of litter flammability in
mixed conifer forests are based largely on burning dry litter.

@@

Field sites for this study were located in Sequoia and Kings Canyon National
Parks, California, USA (N$36^\circ 36'$, W$118^\circ 42'$) between 1600 and
2400\,m elevation in mixed-conifer forest. We chose eight dominant tree species
that are representative of this type of forest: \emph{Pinus jeffreyi} Grev. \&
Balf., \emph{Pinus lambertiana} Dougl., \emph{Pinus ponderosa} Dougl. ex Laws.,
\emph{Abies concolor} (Gord. \& Glend.) Lindl. ex Hildebr., \emph{Abies
  magnifica} A. Murr., \emph{Calocedrus decurrens} (Torr.) Florin,
\emph{Quercus kelloggii} Newb., and \emph{Sequoiadendron giganteum} (Lindl.) J.
Buchholz. In this study, we consider litter the top layer of leaves and small
twigs less than 1\,cm diameter that have fallen in the previous year (mostly
undecomposed or lightly decomposed litter representing fuels classified as
1-hour and 10-hour fuels). Leaf litter was collected in the summer (mid-June to
mid-July) in 2012 from 21 separate sites across the study area. We collected
from a minimum of four populations per species with each population within a
species was 9--32\,km distant from its neighbors. An exception to this was
\emph{A. magnifica} which is restricted to the higher elevations and could only
be collected from one area due to designated wilderness boundaries and
logistical constraints. Within that area, \emph{A. magnifica} litter was
collected from over 10 different individuals. For the remaining species, the
collection involved 2--4 individual trees at each site which were separated
from one another by at least 10\,m. The litter was collected 2\,m away from the
trunk to obtain a more uniform sample of leaf and fine branchlet litter,
because bark and heavier branches tend to fall closer to the tree trunk. The
top 2.5-5\,cm of litter were collected by raking and shoveling into large bags.
All litter was oven dried to $< 5$\% fuel moisture by dry weight, then weighed
and divided into moisture trials. The collected litter was stored in large
paper bags and cardboard boxes in a lab storage space. We minimized handling
and moving the bagged litter to avoid mechanical damage and changes to the
particle size distributions. The moisture manipulations and subsequent burning
trials were conducted in April--August 2015.

@@

We produced dry-down curves to determine maximum water absorption and the rate
of drying for each species and for four distinct three-species mixtures. These
curves allowed us to determine the rate at which a species loses moisture,
approximated as exponential decay, and to determine maximum water content
following saturation and gravity draining. Samples from the eight species (six
replicates per species) and four mixtures (5 replicates per mixture type) were
placed in baskets approximately 45 x 45\,cm by 15\,cm tall constructed of
aluminum mesh with 0.52\,cm square openings. Samples were 450\,g dry weight
litter per trial and litter depth was standardized to 10\,cm by changing the
horizontal dimensions of the basket. Weight was measured by using a balance
sensitive to 0.1\,g (model XS16001L, Mettler Toledo, Columbus, OH). Samples
were immersed in plastic storage bins filled with water for 24 hours to fully
saturate them. After saturation, the baskets were removed from the water,
allowed to drain for 3 minutes, and samples weighed to measure maximum moisture
content. The samples were then allowed to dry in a controlled environment kept
at $21^\circ$C temperature and 30\% relative humidity, assessed via an iButton
sensor (Maxim Integrated) placed 1\,m above the samples. Litter baskets were
arranged in a completely randomized design. During the dry-down process three
subsamples of 5--10\,g were taken at 24 hour intervals and weighed. These
subsamples were then oven dried for 24 hours at $100^\circ$C and re-weighed to
assess moisture content. This was repeated until the final sample from a basket
was less than 10\% moisture. This experimental setup was designed to accelerate
drying compared with field conditions because the baskets were suspended in air
2.8 -- 5.3$^\circ$C above the warmest month average temperatures for the
closest weather stations to the collection sites (Grant Grove and Lodegpole
stations, NOAA 1981-2010 normals,
https://www.ncdc.noaa.gov/cdo-web/datatools/normals).

@@

Burn trials occurred in a cement structure used to simulate house fires at the
Fire Department of the City of Lubbock. This structure minimized wind and
helped regulate temperature and relative humidity. Temperature and relative
humidity were measured before every trial using a Kestrel 3000
(Nielsen-Kellerman, Boothwyn, PA) and these values were used to calculate vapor
pressure deficit as a single weather covariate at each trial. All trials were
conducted on clear days between 10 am and 3 pm. We scheduled burns in a
randomized blocked design aiming to burn one replicate of each species or
mixture on single burn day. This required scheduling species saturation and
during start based on the dry down curves.

Flammability was assessed using a 150\,cm long burn table, 15\,cm wide and
15\,cm tall, in which leaf litter was placed and the table gently shaken to
allow settling. Burning trials all used 450\,g (dry weight) of litter. For
three-species mixtures, this was 150\,g per species. Therefore, fuel mass and
burning tray length and width were fixed, but fuel depth was allowed to vary.
We recorded the actual fuel depth at three points along the tray to calculate
litter bed density before burning. This design allows fire to reach constant
flame spread rates \citep{Magalhaes+Schwilk-2012}. Two metal rulers equally
spaced along the apparatus allowed visual determination of maximum flame
height. Litter was ignited with a propane torch; maximum adiabatic flame
temperature of 1899$^\circ$C,
\url{http://www.benzomatic.com/products/fuel.aspx}) at one end of the apparatus
and allowed to burn until extinction. If a sample failed to ignite after 30
seconds, no further attempt was made and if three samples of a species at a
particular moisture level failed to ignite, no further attempts were made on
that species/moisture combination and the replicate was allowed to continue
drying until the next scheduled checkpoint. The flammability trials followed
the methodology detailed in \citet{Magalhaes+Schwilk-2012}.

During flammability trials, we recorded time to ignition (s), duration of
flaming combustion (s), flame spread rate (mm/s), calculated by dividing the
length of the burned surface by the time it takes the flaming front to reach
the end of the apparatus, maximal flame height (mm), and fuel consumption (\%).
Some trials resulted only in smoldering combustion; for those trials, a time to
ignition and fuel consumption were recorded, but flame spread rate and flame
height were recorded as zero. Our thermocouple system failed or data was lost
for two days of burning trials, therefore we have omitted those temperature
data from this analysis. Our past work \citep{Magalhaes+Schwilk-2012}
demonstrated that flame spread rate captures one important axis of flammability
variation \citep{Schwilk-2015, Pausas+Keeley+etal-2017, Prior+Murphy+etal-2018}
and percent fuel consumed captures a portion of the second major axis of
flammability variation. Our analysis, therefore, focused on these two
relatively orthogonal measures of flammability to avoid examining multiple
redundant dependent variables.

@@

Our past work \citep{Magalhaes+Schwilk-2012} demonstrated that dry litter
flammability across these eight species was driven by leaf traits and litter
packing. These previously reported flammability trials demonstrated that
species within a genus tended to behave similarly and the two scale-leaved
genera in the Cupressaceae, \emph{Calocedrus} and \emph{Sequoiadendron}, also
behaved similarly to one another. Therefore, to preserve power we used
taxonomic groups (\emph{Abies}, Cupressaceae, \emph{Quercus}, and \emph{Pinus})
as our flammability functional groups and treated these as fixed effects in our
models. The effect of fuel moisture and time since wetting on flammability
parameters were modeled using linear models with fuel moisture or time since
wetting and taxon as fixed effects and average vapor pressure deficit in the
burning room during that trial as a nuisance covariate. We used linear mixed
models to predict spread rate, but fuel consumption tends to exhibit threshold
effects with few intermediate values. Therefore, we used mixed generalized
linear models with a beta regression link with the `glmmADMB' package
\citep{Skaug_Fournier_etal-2016} to model percent fuel consumed as a function
of moisture content and of time since wetting.

We investigated possible nonadditive effects on moisture and on flammability by
testing if moisture content and flammability of mixtures were predicted by the
average of the constituent species. To choose mixture combination, we first
grouped the species according to their maximum water absorption and drying rate
according to estimated marginal means using the R package `emmeans'
\citep{Lenth-2019}. This resulted in four groups of species and we then
selected one representative species from each litter type group and produced
all four possible three-species litter mixtures using those four species. For
each mixture and time since wetting, we predicted moisture content based on the
mean moisture content of the three constituent species at that time since
wetting. We predicted flammability parameters based on the average flammability
measures from individual species trials at each time since wetting and then
calculated a predicted spread rate and fuel consumption as the average of these
three values for the mixture. The expectation from the null model was that the
difference between observed and predicted values of flammability will be zero.
We fit linear mixed effect models with the observed-predicted flammability
value as the dependent variable and used Wald tests to test if the centered
intercept was different than zero (a measure of overall nonadditivity) and if
the slope with time since wetting was different from zero (indicating
nonadditivity changing with drying).

@@

\subsection*{Moisture dynamics across species}

Past work has demonstrated that green leaf length predicts litter packing and,
consequently, flame spread rate across these species. Litter particle size
distributions, however show that there is substantial variation in particle
sizes within a species (SI Fig.~1). There were differences in the maximum water
absorption (all $P < 0.0001$, Fig.~\ref{fig:drydown} and
Table~\ref{tab:drydown}). \emph{Q. kelloggii}, the only broadleaf, is the
species that absorbed the most moisture initially. \emph{Quercus}, like the
pines, had a rapid drying rate (Fig.~\ref{fig:drydown} and SI Fig.~2).
Calculating estimated marginal means following the linear mixed-effects model
discriminated the species into three different maximum water absorption groups:
1) \emph{Q. kelloggii}; 2) \emph{P. lambertiana} and \emph{C. decurrens}; and
3) \emph{A. magnifica}, \emph{S. giganteum}, \emph{A. concolor}, \emph{P.
  jeffreyi} and \emph{P. ponderosa} (SI Fig.~2). There are also differences in
drying rates across the eight species and estimated mixed linear model slopes
discriminated species into three groups from lowest to most rapid: 1) \emph{C.
  decurrens}, 2) \emph{Abies spp} and \emph{S. giganteum}, and 3) \emph{Q.
  kelloggii} and \emph{Pinus spp} (SI Fig.~3).

\begin{figure}[h]
  \centering
\includegraphics[width=5cm]{figs_tables/fig1_dry_effects_by_taxon.pdf}
\caption{Effects of litter drying on fuel moisture and flammability. A) Dry
  down curves for eight species. Lines show best-fit exponential decay curves
  fit to moisture content on a dry mass basis. B) Flame spread rate by time
  since wetting. Lines indicate best fit linear model fits by taxonomic group.
  Model ANOVA table in Table \ref{tab:spreadrate_anova} and coefficients in SI
  Table 6. C) Percent fuel consumed by time since wetting. Curves shown
  indicate best fit beta regression lines by taxonomic group. Model ANOVA table
  in Table \ref{tab:consume_anova} and coefficients in SI Table 7. For all
  plots, horizontal jitter (0.5 to 2 hr) was added to avoid overplotting. A
  lack of data at earlier time points for a taxon indicates no ignitions
  occurred at those time points. }
 \label{fig:drydown}
\end{figure}

\begin{table}[h]
  \caption{Linear mixed model results for moisture content as a function of
    time since wetting and species. Approximate degrees of freedom (df), pseudo F
    statistics and p-values were calculated by the Kenward-Roger approximation.
    \citep{Kenward_Roger-1997}. Model coefficients reported in SI Table~1.}
  \label{tab:drydown}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
\begin{tabular}{llll}
  \toprule
Effect & df & $F$ & $P$ \\ 
  \midrule
  hour & 1, 183.73 & 1599.59 & $<$.0001 \\ 
  species & 7, 89.73 & 36.15 & $<$.0001 \\ 
  hour:species & 7, 184.15 & 23.06 & $<$.0001 \\ 
   \bottomrule
\end{tabular}
\end{table}

Across these eight species specific leaf area had a positive association with
maximum water absorption (Table~\ref{tab:mc_di_anova}, SI Fig~5, linear model
$P = 0.003$) and litter bulk density had a negative effect on drying rate
(Table~\ref{tab:mc_di_anova}, SI Fig~6, linear model $P = 0.029$).

\begin{table}[h]
  \caption{Linear model results for maximum water absorption (\% moisture
    content on dry mass basis) and for estimated drying rate (hr$^{-1}$) as
    functions of specific leaf area and litter bulk density across eight
    species. ANOVA tables shown with degrees of freedom, sum of squares, mean
    squares, $F$, and $P$ for each effect. Model coefficients in SI Tables~2 and 3}
  \label{tab:mc_di_anova}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.

>>>>

\documentclass[letterpaper,12pt]{article}
\usepackage{caption}
\DeclareCaptionLabelFormat{adja-page}{\hrulefill\\#1 #2 \emph{(previous page)}}
\usepackage{jecol}
\usepackage{endfloat}

@@

      \item To fill these gaps in knowledge, we determined moisture absorption
    and retention characteristics of the litter from multiple tree species,
    investigated what litter traits influenced these processes, and tested for
    nonadditive moisture dynamics and nonadditive flammability in multi-species
    mixtures. The litter from eight tree species of a mixed-conifer forest was
    saturated and allowed to dry to determine moisture absorption capacity and
    drying rates. Burn trials were performed in litter beds of single species
    and of mixtures to establish flammability response to the dry-down process
    and to species interactions.

      \item Litter mixtures exhibited nonadditivity in flame spread rate and in
    fuel consumption. This nonadditivity became more positive with fuel drying

      \item Moisture dynamics varied across species with higher specific leaf
    area being associated with higher maximum absorption and lower litter bulk
    density being associated with faster drying rates. We found that these
    differing moisture dynamics resulted in time since wetting having strongly
    differing effects on flammability across the major litter types typified by
    different tree taxa..

@@

Fuel moisture content is an important, and often driving, determinant of
surface fire behavior \citep{Rothermel-1972}. Moisture acts as a heat-sink
resulting in less energy available for propagation, and the resulting water
vapor will both cool the flaming front and dilute flammable gases
\citep{Albini-1976, Shafizadeh-1977}. Environmental conditions such as
temperature, relative humidity, and wind drive the moisture content of the fuel
available to burn \citep{Kreye+Varner+etal-2018}. Variation in moisture of
living fuels has strong effects on flammability, but moisture of dead excised
litter fuels can also vary. In particular, fine dead fuels respond readily
to environmental changes \citep{Nelson-2001}. Moisture of litter fuels varies
with climate, with seasons, and with short term weather within a season.
Therefore, both the rate of drying and flammability as a function of moisture
content are key parameters that influence fuel flammability throughout the fire
season.

Flammability is a multi-dimensional characteristic of fuels
\citep{Schwilk-2015, Pausas+Keeley+etal-2017} with different traits influencing
different components of fire behavior. In many forests that experience surface
fire, litter properties influence fire behavior
\citep{Ganteaume+Marielle+etal-2011, Schwilk+Caprio-2011,
  Varner+Kane+etal-2015}. In litter driven fire, the two major axes of
variation are described by measurements related to flame spread rate and
measurements related to heat release or heat total transfer
\citep{Magalhaes+Schwilk-2012, Prior+Murphy+etal-2018}. These different
components of fire behavior have differing ecological effects with spread rate
being associated with flame heights and area burned, and with total heat
transfer predicting soil impacts of fire and tree mortality. A shortcoming in
our current knowledge is our integration of specific species and plant traits
into fuel hazard assessment, fire modeling, and fire effects prediction.
Developing quantitative yet robust and general ways to integrate actual plant
communities into fuel assessment will require understanding the magnitude of
flammability variability across species in space and time and parameterizing
based on plant characteristics.

@@

Fuel moisture content through time is controlled by two processes, wetting
(including absorption and adsorption) and drying rate (desorbtion)
\citep{Anderson-1990, Kreye+Kobziar+etal-2013}. Absorptive capacity represents
the potential maximum fuel moisture and is evaluated by the moisture content of
litter when saturated; hereafter we refer to this measurement as ``maximum
water absorption''. Differences in moisture content over time since saturation
can indicate different abilities to retain moisture. Leaf morphology,
especially leaf cuticle thickness, can influence water absorption
\citep{Van_Wagner-1969}. Litter particle geometry may have effects on maximum
moisture content because a larger total surface area will increase water
adsorption. Drying rate should also be influenced by leaf and litter bed traits
because surface-area-to-volume ratio and litter bulk density will influence air
flow through the litter bed. Therefore, we expect litter packing to effect fire
behavior in two ways: directly by its effect on oxygen availability for
combustion, and indirectly by its effect on drying rate. Both these pathways
are influenced by leaf particle traits, especially leaf size
\citep{Scarff+Westoby-2006}. Fuel-beds comprising larger leaves or leaves with
thin cuticles should favor moisture absorption. Conversely, large leaves lower
litter density and should favor drying, thus increasing the probability of
ignition and fire spread. The interaction of these processes, especially in
species mixtures, is poorly known.

In mixed species communities, species traits may interact such that the
flammability of multi-species mixtures is not the average of the constituent
species. Such nonadditive effects are often synergistic in dry litter fuels
where the flammability of a mixture is driven by the most flammable species
\citep{VanAltena+Logtestjin+etal-2012, Magalhaes+Schwilk-2012}, but negative
nonadditivity has also been recorded for litter mixtures
\citep{Blauw+Wensink+etal-2015, Zhao+vanLogtestijn+etal-2019}. It is possible
that similar nonadditive effects may influence litter moisture content and
flammability of moist fuels. There is some evidence for nonadditive
flammability in moist fuels with a single study finding increasing nonadditivity
with moisture \citep{Blauw+Wensink+etal-2015}, but this previous work did not
include estimation of the dry-down process and had an unclear relationship to
fire behavior in the field. In mixed conifer forest in California, where the
fire regime is characterized by surface fires fueled by plant litter, overstory
tree composition drives variation in local fire behavior
\citep{Schwilk+Caprio-2011} as a result of differential litter flammability
across tree species and as a result of synergistic interactions in
multi-species litter mixtures \citep{Magalhaes+Schwilk-2012}. These
flammability feedbacks may help explain historical fire frequency patterns
\citep{Schwilk+Caprio-2011}, but current estimates of litter flammability in
mixed conifer forests are based largely on burning dry litter.

@@

Field sites for this study were located in Sequoia and Kings Canyon National
Parks, California, USA (N$36^\circ 36'$, W$118^\circ 42'$) between 1600 and
2400\,m elevation in mixed-conifer forest. We chose eight dominant tree species
that are representative of this type of forest: \emph{Pinus jeffreyi} Grev. \&
Balf., \emph{Pinus lambertiana} Dougl., \emph{Pinus ponderosa} Dougl. ex Laws.,
\emph{Abies concolor} (Gord. \& Glend.) Lindl. ex Hildebr., \emph{Abies
  magnifica} A. Murr., \emph{Calocedrus decurrens} (Torr.) Florin,
\emph{Quercus kelloggii} Newb., and \emph{Sequoiadendron giganteum} (Lindl.) J.
Buchholz. In this study, we consider litter the top layer of leaves and small
twigs less than 1\,cm diameter that have fallen in the previous year (mostly
undecomposed or lightly decomposed litter representing fuels classified as
1-hour and 10-hour fuels). Leaf litter was collected in the summer (mid-June to
mid-July) in 2012 from 21 separate sites across the study area. We collected
from a minimum of four populations per species with each population within a
species was 9--32\,km distant from its neighbors. An exception to this was
\emph{A. magnifica} which is restricted to the higher elevations and could only
be collected from one area due to designated wilderness boundaries and
logistical constraints. Within that area, \emph{A. magnifica} litter was
collected from over 10 different individuals. For the remaining species, the
collection involved 2--4 individual trees at each site which were separated
from one another by at least 10\,m. The litter was collected 2\,m away from the
trunk to obtain a more uniform sample of leaf and fine branchlet litter,
because bark and heavier branches tend to fall closer to the tree trunk. The
top 2.5-5\,cm of litter were collected by raking and shoveling into large bags.
All litter was oven dried at $60^\circ$C to $< 5$\% fuel moisture by dry
weight, then weighed and divided into moisture trials. The collected litter was
stored in large paper bags and cardboard boxes in a lab storage space. We
minimized handling and moving the bagged litter to avoid mechanical damage and
changes to the particle size distributions. The moisture manipulations and
subsequent burning trials were conducted in April--August 2015.

@@

We produced dry-down curves to determine maximum water absorption and the rate
of drying for each species and for four distinct three-species mixtures. These
curves allowed us to determine the rate at which a species loses moisture,
approximated as exponential decay, and to determine maximum water content
following saturation and gravity draining. Samples from the eight species (six
replicates per species) and four mixtures (five replicates per mixture type) were
placed in baskets approximately 45 x 45\,cm by 15\,cm tall constructed of
aluminum mesh with 0.52\,cm square openings. Samples were 450\,g dry weight
litter per trial and litter depth was standardized to 10\,cm by changing the
horizontal dimensions of the basket. Weight was measured by using a balance
sensitive to 0.1\,g (model XS16001L, Mettler Toledo, Columbus, OH). Samples
were immersed in plastic storage bins filled with water for 24 hours to fully
saturate them. After saturation, the baskets were removed from the water,
allowed to drain for 3 minutes, and samples weighed to measure maximum moisture
content. The samples were then allowed to dry in a controlled environment kept
at $21^\circ$C temperature and 30\% relative humidity, assessed via an iButton
sensor (Maxim Integrated) placed 1\,m above the samples. Litter baskets were
arranged in a completely randomized design. During the dry-down process three
subsamples of 5--10\,g were taken at 24 hour intervals and weighed. These
subsamples were then oven dried for 24 hours at $100^\circ$C and re-weighed to
assess moisture content. This was repeated until the final sample from a basket
was less than 10\% moisture. This experimental setup was designed to accelerate
drying compared with field conditions because the baskets were suspended in air that was
2.8 -- 5.3$^\circ$C above the warmest month average temperatures for the
closest weather stations to the collection sites (Grant Grove and Lodegpole
stations, NOAA 1981-2010 normals,
https://www.ncdc.noaa.gov/cdo-web/datatools/normals).

@@

Burn trials occurred in a cement structure used to simulate house fires at the
Fire Department of the City of Lubbock. This structure minimized wind and
helped regulate temperature and relative humidity. Temperature and relative
humidity were measured before every trial using a Kestrel 3000
(Nielsen-Kellerman, Boothwyn, PA) and these values were used to calculate vapor
pressure deficit as a weather covariate at each trial. All trials were
conducted on clear days between 10 am and 3 pm. We scheduled burns in a
randomized blocked design aiming to burn one replicate of each species or
mixture on single burn day. This required scheduling species saturation and
during start based on the dry down curves.

Flammability was assessed using a 150\,cm long burn table, 15\,cm wide and
15\,cm tall, in which leaf litter was placed and the table gently shaken to
allow settling. Burning trials all used 450\,g (dry weight) of litter. For
three-species mixtures, this was 150\,g per species. Therefore, fuel mass and
burning tray length and width were fixed, but fuel depth was allowed to vary.
We recorded the actual fuel depth at three points along the tray to calculate
litter bed density before burning. This design allows fire to reach constant
flame spread rates within the range of rates reported for field burns \citep{Magalhaes+Schwilk-2012}. Two metal rulers equally
spaced along the apparatus allowed visual determination of maximum flame
height. Litter was ignited with a propane torch; maximum adiabatic flame
temperature of 1899$^\circ$C,
\url{http://www.benzomatic.com/products/fuel.aspx}) at one end of the apparatus
and allowed to burn until extinction. If a sample failed to ignite after 30
seconds, no further attempt was made and if three samples of a species at a
particular moisture level failed to ignite, no further attempts were made on
that species/moisture combination and the replicate was allowed to continue
drying until the next scheduled checkpoint. The flammability trials followed
the methodology detailed in \citet{Magalhaes+Schwilk-2012}.

During flammability trials, we recorded time to ignition (s), duration of
flaming combustion (s), flame spread rate (mm/s), calculated by dividing the
length of the burned surface by the time it takes the flaming front to reach
the end of the apparatus, maximal flame height (mm), and fuel consumption (\%).
Some trials resulted only in smoldering combustion; for those trials, a time to
ignition and fuel consumption were recorded, but flame spread rate and flame
height were recorded as zero. Our thermocouple system failed or data was lost
for two days of burning trials, therefore we have omitted temperature
data from this analysis. Our past work \citep{Magalhaes+Schwilk-2012}
demonstrated that flame spread rate captures one important axis of flammability
variation \citep{Schwilk-2015, Pausas+Keeley+etal-2017, Prior+Murphy+etal-2018}
and percent fuel consumed captures a portion of the second major axis of
flammability variation. Our analysis, therefore, focused on these two
relatively orthogonal measures of flammability to avoid examining multiple
redundant dependent variables.

@@

Our past work \citep{Magalhaes+Schwilk-2012} demonstrated that dry litter
flammability across these eight species was driven by leaf traits and litter
packing. These previously reported flammability trials demonstrated that
species within a genus tended to behave similarly and that the two scale-leaved
genera in the Cupressaceae, \emph{Calocedrus} and \emph{Sequoiadendron}, also
behaved similarly to one another. Therefore, to preserve power we used
taxonomic groups (\emph{Abies}, Cupressaceae, \emph{Quercus}, and \emph{Pinus})
as our flammability functional groups and treated these as fixed effects in our
models. The effects of fuel moisture and time since wetting on flammability
parameters were modeled using linear models with fuel moisture or time since
wetting and taxon as fixed effects and average vapor pressure deficit in the
burning room during that trial as a nuisance covariate. We used linear mixed
models to predict spread rate, but fuel consumption tends to exhibit threshold
effects with few intermediate values. Therefore, we used mixed generalized
linear models with a beta regression link with the `glmmADMB' package
\citep{Skaug_Fournier_etal-2016} to model percent fuel consumed as a function
of moisture content and of time since wetting.

We investigated nonadditive effects on moisture and on flammability by
testing if moisture content and flammability of mixtures were predicted by the
average of the constituent species. To choose mixture combinations, we first
grouped the species according to their maximum water absorption and drying rate
according to estimated marginal means using the R package `emmeans'
\citep{Lenth-2019}. This resulted in four groups of species and we then
selected one representative species from each litter type group and produced
all four possible three-species litter mixtures using those four species. For
each mixture and time since wetting, we predicted moisture content based on the
mean moisture content of the three constituent species at that time since
wetting. We predicted flammability parameters based on the average flammability
measures from individual species trials at each time since wetting and then
calculated a predicted spread rate and fuel consumption as the average of these
three values for the mixture. The expectation from the null model was that the
difference between observed and predicted values of flammability will be zero.
We fit linear mixed effect models with the observed-predicted flammability
value as the dependent variable and used Wald tests to test if the centered
intercept was different than zero (a measure of overall nonadditivity) and if
the slope with time since wetting was different from zero (indicating
nonadditivity changing with drying).

@@

\subsection*{Moisture dynamics}

Past work has demonstrated that green leaf length predicts litter packing and,
consequently, flame spread rate across these species. Litter particle size
distributions, however show that there is substantial variation in particle
sizes within a species (SI Fig.~1). There were differences in the maximum water
absorption (all $P < 0.0001$, Fig.~\ref{fig:drydown} and
Table~\ref{tab:drydown}). \emph{Q. kelloggii}, the only broadleaf, is the
species that absorbed the most moisture initially. \emph{Quercus}, like the
pines, had a rapid drying rate (Fig.~\ref{fig:drydown} and SI Fig.~2).
Calculating estimated marginal means following the linear mixed-effects model
discriminated the species into three different maximum water absorption groups:
1) \emph{Q. kelloggii}; 2) \emph{P. lambertiana} and \emph{C. decurrens}; and
3) \emph{A. magnifica}, \emph{S. giganteum}, \emph{A. concolor}, \emph{P.
  jeffreyi} and \emph{P. ponderosa} (SI Fig.~2). There are also differences in
drying rates across the eight species and estimated mixed linear model slopes
discriminated species into three groups from lowest to most rapid: 1) \emph{C.
  decurrens}, 2) \emph{Abies spp.} and \emph{S. giganteum}, and 3) \emph{Q.
  kelloggii} and \emph{Pinus spp.} (SI Fig.~3).

\begin{figure}
  \centering
  \includegraphics[width=8cm]{figs_tables/fig1_dry_effects_by_taxon.pdf}
  \caption{}
\end{figure}
%%\clearpage
\begin{figure}
  \captionsetup{labelformat=adja-page}
  \ContinuedFloat
\caption{Effects of litter drying on fuel moisture and flammability. A) Dry
  down curves for eight species. Lines show best-fit exponential decay curves
  fit to moisture content on a dry mass basis. B) Flame spread rate by time
  since wetting. Lines indicate best fit linear model fits by taxonomic group.
  Model ANOVA table in Table \ref{tab:spreadrate_anova} and coefficients in SI
  Table 6. C) Percent fuel consumed by time since wetting. Curves shown
  indicate best fit beta regression lines by taxonomic group. Model ANOVA table
  in Table \ref{tab:consume_anova} and coefficients in SI Table 7. For all
  plots, horizontal jitter (0.5 to 2 hr) was added to avoid overplotting. A
  lack of data at earlier time points for a taxon indicates no ignitions
  occurred at those time points. }
 \label{fig:drydown}
\end{figure}
%% \clearpage

\begin{table}
  \caption{Linear mixed model results for moisture content as a function of
    time since wetting and species. Approximate degrees of freedom (df), pseudo F
    statistics and p-values were calculated by the Kenward-Roger approximation.
    \citep{Kenward_Roger-1997}. Model coefficients are reported in SI Table~1.}
  \label{tab:drydown}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
\begin{tabular}{llll}
  \toprule
Effect & df & $F$ & $P$ \\ 
  \midrule
  hour & 1, 183.73 & 1599.59 & $<$.0001 \\ 
  species & 7, 89.73 & 36.15 & $<$.0001 \\ 
  hour:species & 7, 184.15 & 23.06 & $<$.0001 \\ 
   \bottomrule
\end{tabular}
\end{table}

Across these eight species, specific leaf area had a positive association with
maximum water absorption (Table~\ref{tab:mc_di_anova}, SI Fig~5, linear model
$P = 0.003$) and litter bulk density had a negative effect on drying rate
(Table~\ref{tab:mc_di_anova}, SI Fig~6, linear model $P = 0.029$).
